---
title: "Background Model on Newsgroups Dataset"
author: "Jihee You"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE
)
library(here)
library(Matrix)
library(ggplot2)
library(cowplot)
library(fastTopics)
source(here("docs", "code", "bg_implement.R"), local = TRUE)
source(here("docs", "code", "bg_post.R"), local = TRUE)
```

# Experiments on Newsgroups

Both models use the full document-by-term count matrix by default.

```{r load}
news <- load_news_data(
  here("docs", "data", "newsgroups.RData"),
  here("docs", "data", "20news-bydate", "train.map")
)
```

```{r run}
K <- 10L
a <- c(0.1, 1, 10, 100, 1000)
ebpmf_maxiter <- 600L
fasttopics_numiter <- 600L
X <- news$X

set.seed(1)

stopifnot(nrow(X) > 0L, ncol(X) > 0L, all(Matrix::colSums(X) > 0))
ebpmf_elapsed <- system.time({
  state <- initialize_state(X, K, a, L_init = "deviance_kmeans")
  initial_state <- state
  ebpmf <- run_ebpmf_bg_cavi(
    state, X, a, max_niter = ebpmf_maxiter, early_stop = FALSE
  )
})[["elapsed"]]

fasttopics_elapsed <- system.time({
  fasttopics <- fastTopics::fit_poisson_nmf(
    X, k = K, numiter = fasttopics_numiter,
    init.method = "random", verbose = "none"
  )
})[["elapsed"]]
fits <- list(
  ebpmf = ebpmf,
  fasttopics = fasttopics,
  dimensions = data.frame(documents = nrow(X), terms = ncol(X), topics = K),
  timing = data.frame(
    method = c("EBPMF-BG", "fastTopics"),
    initialization = c("deviance k-means L; all-one F", "random"),
    iterations = c(ebpmf$completed_niter, fasttopics_numiter),
    elapsed_seconds = c(ebpmf_elapsed, fasttopics_elapsed)
  )
)
fits$dimensions
fits$timing
```

Both models are fitted to the same sparse matrix. Timings include initialization and fitting.

### See Initialized Loadings

Document-topic proportions from the initial background model loadings, before CAVI updates.

```{r initial-L, fig.width=14, fig.height=4, dpi=150}
plot_news_structure(initial_state$L, news$labels)
```

### Align the fitted topics

```{r newsgroups-align}
aligned <- align_news_models(fits)
aligned$alignment_table
```

The assignment matches fastTopics to EBPMF using the average cosine similarity of document-topic proportions and word profiles. 

For EBPMF topic k and fastTopics topic h, the matching score is
$$
S_{kh}=
\frac{
\operatorname{cosine}(\text{doc-topic-prop}^{E}_{\cdot k},\text{doc-topic-prop}^{F}_{\cdot h})
+
\operatorname{cosine}(F^{E}_{\cdot k},F^{F}_{\cdot h})
}{2}.
$$
Topics match well when they have similar documents (in terms of topic contribution degree) and similar words.


### Convergence diagnostics

```{r newsgroups-convergence, fig.width=12, fig.height=8, dpi=150}
plot_news_convergence(fits)
```

The fastTopics likelihood and EBPMF ELBO are different objectives; their
magnitudes are not directly comparable.


### Plot mixprop estimations

```{r newsgroups-mixprop, fig.width=7, fig.height=6}
plot_news_mixprop(fits$ebpmf$state$pi, a)
```


### Loading and word-frequency agreement

```{r newsgroups-factor-agreement, fig.width=14, fig.height=10, dpi=150}
plot_news_agreement(aligned, news$terms)
```


### Document-topic structure

```{r newsgroups-structure, fig.width=14, fig.height=7.5, dpi=150}
# The same horizontal position represents different documents in the two panels.
plot_news_structure(aligned, news$labels, plot_sample = TRUE)
plot_news_structure(aligned, news$labels, plot_sample = FALSE)
```

Both panels show the same sampled documents (sampled after full dataset fitting), grouped by observed newsgroup.
Each model uses its own seeded 1D t-SNE ordering within each newsgroup for a smoother document topic graph. Therefore, the horizontal position does not represent same document in the two panels. 

### Keywords

Top words in each EBPMF topic, ranked by raw fitted $F_{jk}$.
For EBPMF, these are posterior means $E_q[F_{jk}]$, without the background
multiplier $f_{0j}$. These values are not normalized topic-word probabilities.

```{r newsgroups-top-words}
# Set n to 5, 10, 20, etc.; also lists recurring words and their topics.
news_top_words(fits, news$terms, n = 5)
```

### Investigating a keyword

Change `word`, `topic` below.

```{r keyword-controls}
word <- "stephanopoulos"
topic <- 6L
```

Summary object returns more specific occurrences of a certain word. On the other hand, document object prints information of a single document containing the word, including the sorted word count of one document containing the word. 

Total_tokens 5044 means there were total of 5044 words in that document, whereas word_count is specific to the word of interest. topic_proportion means every document containing the selected word has at the lowest/highest, such ratio of document-topic proportions related to the selected topic.

```{r keyword-documents}
# Return n documents per group: highest and lowest topic membership.
news_word_documents(news, aligned$L_ebpmf_topics, word = word, k = topic, n=1) 
```

The `membership` column identifies the highest and lowest groups. Both groups
contain only documents with the searched word; they may overlap when few
documents contain it.

### Topic sizes

`effective_documents` sums document-topic proportions. `dominant_documents`
assigns each document to the topic associated with its largest document-topic proportion.

```{r topic-sizes}
news_topic_sizes(news, aligned$L_ebpmf_topics)
```

<!--
### Keyword-like topic

```{r newsgroups-keyword-topic, fig.width=12, fig.height=5.5, dpi=150}
plot_news_keyword(aligned, news$X, news$terms)
```

The selected topic places the most average mass on words occurring across many
documents. Labels identify words enriched relative to other topics.
-->

### Distinctive words

```{r newsgroups-distinct}
news_distinctive_words(aligned, news$terms)
```

A different notion of keywords. Words that do not appear frequently in other topics but appear in that topic.


### Note on scaling
Simply normalizing the loadings do not make sense. For example,
If a document has $L_i = (2,2)$ simple normalization would yield $(0.5,0.5)$ document-topic proportion, but the two topics could have very different total feature masses $\sum_j F_{j1} = 10, \sum_j F_{j2} = 2$. Accounting for this, topic 1's contribution to the document i's expected total word count is $\frac{2*10}{24}$, and topic 2's contribution is $\frac{2*2}/24$.
This is the document-topic proportion we are interested in. Note for the bg model, $F_{jk}$ would be replaced with $f_0 F_{jk}$. 
$$
\frac{
L_{ik}\sum_jF_{jk}
}{
\sum_h L_{ih}\sum_jF_{jh}
}
$$

### Note on background, keywords
In some contexts, the background model makes more sense than a sparse model because . And how this model might naturally lead to identification of key genes or key words? Higher $F_{jk}$ means more 

Where keywords are associated with smaller a,b and background words are associated with larger a,b (with larger variance).
