---
title: "bg_news_badsample"
author: "Jihee You"
date: "`r Sys.Date()`"
output: html_document
---
---
title: "Background Model on Newsgroups Dataset"
author: "Jihee You"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE
)
library(here)
library(Matrix)
library(ggplot2)
library(cowplot)
library(fastTopics)
source(here("docs", "code", "bg_implement.R"), local = TRUE)
source(here("docs", "code", "bg_post.R"), local = TRUE)
```

# Experiments on Newsgroups

Both models use the full document-by-term count matrix by default.

```{r load}
news <- load_news_data(
  here("docs", "data", "newsgroups.RData"),
  here("docs", "data", "20news-bydate", "train.map")
)
```

However, the optional `sample` chunk selects random documents and the most frequent words, then removes words absent from those documents.

```{r sample}
news <- sample_news_data(news, max_documents = 1000L, max_terms = 2000L)
```

```{r run}
K <- 10L
a <- c(0.1, 1, 10, 100, 1000)
ebpmf_maxiter <- 600L
fasttopics_numiter <- 600L
X <- news$X

set.seed(1)

stopifnot(nrow(X) > 0L, ncol(X) > 0L, all(Matrix::colSums(X) > 0))
ebpmf_elapsed <- system.time({
  state <- initialize_state(X, K, a, L_init = "deviance_kmeans")
  initial_state <- state
  ebpmf <- run_ebpmf_bg_cavi(
    state, X, a, max_niter = ebpmf_maxiter, early_stop = FALSE
  )
})[["elapsed"]]

fasttopics_elapsed <- system.time({
  fasttopics <- fastTopics::fit_poisson_nmf(
    X, k = K, numiter = fasttopics_numiter,
    init.method = "random", verbose = "none"
  )
})[["elapsed"]]
fits <- list(
  ebpmf = ebpmf,
  fasttopics = fasttopics,
  dimensions = data.frame(documents = nrow(X), terms = ncol(X), topics = K),
  timing = data.frame(
    method = c("EBPMF-BG", "fastTopics"),
    initialization = c("deviance k-means L; all-one F", "random"),
    iterations = c(ebpmf$completed_niter, fasttopics_numiter),
    elapsed_seconds = c(ebpmf_elapsed, fasttopics_elapsed)
  )
)
fits$dimensions
fits$timing
```

Both models are fitted to the same sparse matrix. Timings include initialization and fitting.

### See Initialized Loadings

Document-topic proportions from the initial background model loadings, before CAVI updates.

```{r initial-L, fig.width=14, fig.height=4, dpi=150}
plot_news_structure(initial_state$L, news$labels)
```

### Align the fitted topics

```{r newsgroups-align}
aligned <- align_news_models(fits)
aligned$alignment_table
```

The assignment matches fastTopics to EBPMF using the average cosine similarity of document-topic proportions and word profiles. 

For EBPMF topic k and fastTopics topic h, the matching score is
$$
S_{kh}=
\frac{
\operatorname{cosine}(\text{doc-topic-prop}^{E}_{\cdot k},\text{doc-topic-prop}^{F}_{\cdot h})
+
\operatorname{cosine}(F^{E}_{\cdot k},F^{F}_{\cdot h})
}{2}.
$$
Topics match well when they have similar documents (in terms of topic contribution degree) and similar words.


### Convergence diagnostics

```{r newsgroups-convergence, fig.width=12, fig.height=8, dpi=150}
plot_news_convergence(fits)
```

The fastTopics likelihood and EBPMF ELBO are different objectives; their
magnitudes are not directly comparable.

### Document-topic structure

```{r newsgroups-structure, fig.width=14, fig.height=7.5, dpi=150}
# The same horizontal position represents different documents in the two panels.
plot_news_structure(aligned, news$labels, plot_sample = FALSE)
```

Both panels show the same sampled documents (sampled after full dataset fitting), grouped by observed newsgroup.
Each model uses its own seeded 1D t-SNE ordering within each newsgroup for a smoother document topic graph. Therefore, the horizontal position does not represent same document in the two panels.

# With different initialization
We don't see k10 dominate anymore.

```{r}
K <- 10L
a <- c(0.1, 1, 10, 100, 1000)
ebpmf_maxiter <- 500L
fasttopics_numiter <- 500L
X <- news$X

set.seed(1)

stopifnot(nrow(X) > 0L, ncol(X) > 0L, all(Matrix::colSums(X) > 0))
ebpmf_elapsed <- system.time({
  state <- initialize_state(X, K, a, L_init = "fasttopics")
  initial_state <- state
  ebpmf <- run_ebpmf_bg_cavi(
    state, X, a, max_niter = ebpmf_maxiter, early_stop = FALSE
  )
})[["elapsed"]]

fasttopics_elapsed <- system.time({
  fasttopics <- fastTopics::fit_poisson_nmf(
    X, k = K, numiter = fasttopics_numiter,
    init.method = "random", verbose = "none"
  )
})[["elapsed"]]
fits <- list(
  ebpmf = ebpmf,
  fasttopics = fasttopics,
  dimensions = data.frame(documents = nrow(X), terms = ncol(X), topics = K),
  timing = data.frame(
    method = c("EBPMF-BG", "fastTopics"),
    initialization = c("deviance k-means L; all-one F", "random"),
    iterations = c(ebpmf$completed_niter, fasttopics_numiter),
    elapsed_seconds = c(ebpmf_elapsed, fasttopics_elapsed)
  )
)
fits$dimensions
fits$timing
```

Both models are fitted to the same sparse matrix. Timings include initialization and fitting.

### See Initialized Loadings

Document-topic proportions from the initial background model loadings, before CAVI updates.

```{r fig.width=14, fig.height=4, dpi=150}
plot_news_structure(initial_state$L, news$labels)
```

### Align the fitted topics

```{r}
aligned <- align_news_models(fits)
aligned$alignment_table
```

The assignment matches fastTopics to EBPMF using the average cosine similarity of document-topic proportions and word profiles. 

For EBPMF topic k and fastTopics topic h, the matching score is
$$
S_{kh}=
\frac{
\operatorname{cosine}(\text{doc-topic-prop}^{E}_{\cdot k},\text{doc-topic-prop}^{F}_{\cdot h})
+
\operatorname{cosine}(F^{E}_{\cdot k},F^{F}_{\cdot h})
}{2}.
$$
Topics match well when they have similar documents (in terms of topic contribution degree) and similar words.


### Convergence diagnostics

```{r fig.width=12, fig.height=8, dpi=150}
plot_news_convergence(fits)
```

The fastTopics likelihood and EBPMF ELBO are different objectives; their
magnitudes are not directly comparable.

### Document-topic structure

```{r fig.width=14, fig.height=7.5, dpi=150}
# The same horizontal position represents different documents in the two panels.
plot_news_structure(aligned, news$labels, plot_sample = FALSE)
```

Both panels show the same sampled documents (sampled after full dataset fitting), grouped by observed newsgroup.
Each model uses its own seeded 1D t-SNE ordering within each newsgroup for a smoother document topic graph. Therefore, the horizontal position does not represent same document in the two panels.


