---
title: "BG Model on Simulated Data"
author: "Jihee You"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# I want to run something like 
# run_rmd_chunk("analysis/bg_sim.Rmd", "main-loop")
# later

knitr::opts_chunk$set(echo = TRUE)
library(here)
here()

run_rmd_chunk <- function(file, label, envir = globalenv()) {
  lines <- readLines(file)

  start <- grep(
    sprintf("^```\\{r[[:space:]]+%s([,[:space:]]|\\})", label),
    lines
  )

  if (length(start) != 1L) {
    stop("Chunk label not found or is not unique: ", label)
  }

  end <- which(seq_along(lines) > start & lines == "```")[1]

  if (is.na(end)) {
    stop("Closing code fence not found.")
  }

  eval(
    parse(text = lines[(start + 1L):(end - 1L)]),
    envir = envir
  )
}
```

## Code to run experiments on synthetic data

```{r import}
library(here)
here()
source(here("docs","code","bg_implement.R"))
```

Run ebpmf-bg full loop on simulated data.

```{r main-loop-controls}
a_grid <- c(0.1, 1, 10, 100, 1000)
pi_mat <- cbind(
  mostly_background = c(0, 0, 0, 0.05, 0.95),
  broad_signal = c(0, 0.05, 0, 0, 0.95),
  near_zero_signal = c(0.05, 0, 0, 0, 0.95)
)
# Create data
settings <- list(
  n = 300, 
  p = 500, 
  K = 3,
  a = a_grid, # length-M Gamma shape list
  pi = pi_mat,
  seed = 1L,
  loading_mode = "single" # "single", "primary_secondary"
)

synthetic <- do.call(simulate_background_model, settings)


# Experiment switches.
init <- "deviance_kmeans" # "gamma", "deviance_kmeans", or "fasttopics"
oracle_f0 <- FALSE
oracle_pi <- FALSE
oracle_l <- FALSE

# Fit controls.
max_niter <- 400L
elbo_every <- 10L
fit_elbo_tolerance <- 1e-6
fit_elbo_patience <- 3L
```

```{r main-loop}
init <- match.arg(init, c("gamma", "deviance_kmeans", "fasttopics"))

# Avoid an unnecessary k-means run when oracle L will replace it.
state <- initialize_state(
  synthetic$X,
  settings$K,
  settings$a,
  L_init = if (oracle_l) "gamma" else init,
  seed = 2L
)

# Apply any known parameters. Oracle L is initialization only; CAVI updates it.
if (oracle_f0) state$f0 <- synthetic$f0
if (oracle_pi) {
  state$pi <- synthetic$pi
  state$qF <- initialize_qF(state$pi, settings$a, ncol(synthetic$X))
}
if (oracle_l) state$L <- synthetic$L

# Label the initialization used by the diagnostic summary.
initialization_method <- if (oracle_l) "oracle L" else switch(
  init,
  gamma = "Gamma-random L",
  deviance_kmeans = "deviance k-means L",
  fasttopics = "fastTopics L"
)

# Save the exact parameters entering CAVI for the diagnostic plots below.
L_initial <- state$L
F_initial <- state$qF$mean
pi_initial <- state$pi
f0_initial <- state$f0

# run_ebpmf_bg_cavi() synchronizes qZ before evaluating the initial ELBO.
M <- length(settings$a)
main_fit <- run_ebpmf_bg_cavi(
  state,
  synthetic$X,
  settings$a,
  max_niter = max_niter,
  elbo_every = elbo_every,
  fit_elbo_tolerance = fit_elbo_tolerance,
  fit_elbo_patience = fit_elbo_patience,
  estimate_pi = !oracle_pi,
  estimate_f0 = !oracle_f0
)

state <- main_fit$state
elbo_trace <- main_fit$elbo_trace
pi_trace <- main_fit$pi_trace
niter <- main_fit$completed_niter

if (oracle_pi) stopifnot(max(abs(state$pi - synthetic$pi)) < 1e-15)

# Summarize and verify ELBO changes between stored checkpoints.
elbo_comparison <- data.frame(
  iteration = elbo_trace$iteration[-1L],
  elbo = tail(elbo_trace$elbo, -1L),
  change = diff(elbo_trace$elbo),
  relative_change = elbo_trace$relative_change[-1L]
)
stopifnot(all(elbo_comparison$change >= -1e-8))

fit_summary <- data.frame(
  maximum_iterations = max_niter,
  completed_iterations = niter,
  elbo_every = elbo_every,
  converged = main_fit$converged,
  oracle_f0 = oracle_f0,
  oracle_pi = oracle_pi,
  oracle_l = oracle_l,
  initialization = initialization_method
)

tail(elbo_comparison, 10L)
fit_summary

```



## Diagnostic Plots
The plots below provide diagnostic tools for interpreting EBPMF-BG runs. I eventually plan to use them in a coherent experimental log that includes earlier unsuccessful results; for now, this section documents only the visualization code.

### Initial parameter values

The panels below show the state that enters the main CAVI loop. `F_initial` is
the variational mean `qF$mean`. Factor columns are shown in their actual fitting
order, without post-hoc alignment to the simulated truth.

```{r initial-parameters, fig.width=13, fig.height=9, dpi=150}
initial_heatmap <- function(
    value,
    main,
    row_label,
    row_axis_at = NULL,
    row_axis_labels = NULL
) {
  color_limit <- range(value, finite = TRUE)
  if (diff(color_limit) == 0) {
    padding <- max(abs(color_limit[1L]), 1) * 1e-8
    color_limit <- color_limit + c(-padding, padding)
  }

  image(
    x = seq_len(ncol(value)),
    y = seq_len(nrow(value)),
    z = t(value),
    col = hcl.colors(80L, "Viridis"),
    zlim = color_limit,
    xlab = "Factor",
    ylab = row_label,
    main = main,
    axes = FALSE,
    useRaster = TRUE
  )
  axis(1, at = seq_len(ncol(value)), labels = seq_len(ncol(value)))

  if (is.null(row_axis_at)) {
    row_axis_at <- unique(c(1L, nrow(value)))
    row_axis_labels <- row_axis_at
  }
  axis(
    2,
    at = row_axis_at,
    labels = row_axis_labels,
    las = 1,
    cex.axis = 0.8
  )
  box()
  mtext(
    sprintf("range: %.3g to %.3g", min(value), max(value)),
    side = 3,
    line = 0.2,
    cex = 0.75
  )
}

old_par <- par(mfrow = c(2, 2), mar = c(2.5, 3, 2, 0.5))

initial_heatmap(
  L_initial,
  main = "Initial L",
  row_label = "Document"
)

initial_heatmap(
  F_initial,
  main = "Initial E_q[F]",
  row_label = "Feature"
)

initial_heatmap(
  pi_initial,
  main = "Initial pi",
  row_label = "Gamma component",
  row_axis_at = seq_len(nrow(pi_initial)),
  row_axis_labels = paste0("a=", settings$a)
)

plot(
  seq_along(f0_initial),
  f0_initial,
  type = "l",
  lwd = 1.5,
  col = "steelblue",
  xlab = "Feature",
  ylab = "f0",
  main = "Initial f0"
)
abline(h = 1, col = "grey50", lty = 2)
grid(col = "grey90")

par(old_par)
```


Plot the log-scaled gap from the best recorded ELBO. Values decrease toward
`log10(elbo_plot_epsilon)` as the fit converges.

```{r elbo-iteration, fig.width=9, fig.height=5, dpi=150}
elbo_plot_epsilon <- 1e-8
best_elbo <- max(elbo_trace$elbo)
elbo_by_iteration <- head(
  elbo_trace[, c("iteration", "elbo")],
  -1L
)
elbo_by_iteration$log10_gap <- log10(
  best_elbo - elbo_by_iteration$elbo + elbo_plot_epsilon
)

plot(
  log10_gap ~ iteration,
  data = elbo_by_iteration,
  type = "b",
  pch = 19,
  lwd = 2,
  col = "steelblue",
  xlab = "CAVI iteration",
  ylab = expression(log[10](max(ELBO) - ELBO + epsilon)),
  main = "ELBO convergence"
)
grid(col = "grey90")

# Label the final visible checkpoint (iteration 390 in the current run).
last_point <- tail(elbo_by_iteration, 1L)
text(
  last_point$iteration,
  last_point$log10_gap,
  labels = paste("Iteration", last_point$iteration),
  pos = 2
)
```


### Aligned factor recovery for easier analysis

Factor labels and scales are not intrinsically identifiable: components may be
permuted. The comparisons below therefore align
components by their rate contributions (with cosine similarity) and adopt a common scale before
assessing recovery.

```{r align-normalize}
## Posterior point estimates
L_true <- synthetic$L
F_true <- synthetic$F
f0_true <- synthetic$f0
pi_true <- synthetic$pi

L_hat <- state$L
F_hat <- state$qF$mean
f0_hat <- state$f0
pi_hat <- state$pi

K <- ncol(L_true)

## Effective feature factors: these are what enter the likelihood
B_true <- sweep(F_true, 1, f0_true, "*")
B_hat  <- sweep(F_hat,  1, f0_hat,  "*")


## ------------------------------------------------------------------
## 1. Match estimated components to true components
## ------------------------------------------------------------------

# For rank-one matrices, <l1 b1', l2 b2'> = <l1, l2><b1, b2> and
# ||l b'||_F = ||l|| ||b||. Use these identities to avoid constructing any
# dense n-by-p component-specific Poisson-rate matrices.
loading_inner_product <- crossprod(L_true, L_hat)
feature_inner_product <- crossprod(B_true, B_hat)
true_component_norm <-
  sqrt(diag(crossprod(L_true))) * sqrt(diag(crossprod(B_true)))
fitted_component_norm <-
  sqrt(diag(crossprod(L_hat))) * sqrt(diag(crossprod(B_hat)))

component_similarity <-
  (loading_inner_product * feature_inner_product) /
  outer(true_component_norm, fitted_component_norm)

# Base-R permutation generator; practical for a modest number of factors.
all_permutations <- function(x) {
  if (length(x) == 1L) {
    return(matrix(x, nrow = 1L))
  }

  do.call(
    rbind,
    lapply(seq_along(x), function(i) {
      cbind(x[i], all_permutations(x[-i]))
    })
  )
}

permutations <- all_permutations(seq_len(K))

permutation_score <- apply(permutations, 1, function(perm) {
  sum(component_similarity[cbind(seq_len(K), perm)])
})

best_permutation <- permutations[which.max(permutation_score), ]

# Reorder estimated quantities to match the true components.
L_hat <- L_hat[, best_permutation, drop = FALSE]
F_hat <- F_hat[, best_permutation, drop = FALSE]
B_hat <- B_hat[, best_permutation, drop = FALSE]
pi_hat <- pi_hat[, best_permutation, drop = FALSE]


## ------------------------------------------------------------------
## 2. Resolve the remaining component-wise L/B scale ambiguity
## ------------------------------------------------------------------
## L[, k] * B[, k]' is unchanged if L[, k] is divided by c and
## B[, k] is multiplied by c. Use mean(L[, k]) = 1 as a convention.

canonicalize <- function(L, B) {
  component_scale <- colMeans(L)

  list(
    L = sweep(L, 2, component_scale, "/"),
    B = sweep(B, 2, component_scale, "*")
  )
}

true_canonical <- canonicalize(L_true, B_true)
hat_canonical  <- canonicalize(L_hat, B_hat)

L_true_plot <- true_canonical$L
L_hat_plot  <- hat_canonical$L

B_true_plot <- true_canonical$B
B_hat_plot  <- hat_canonical$B


## ------------------------------------------------------------------
## 3. Direct factor-comparison plots
## ------------------------------------------------------------------

plot_comparison <- function(truth, estimate, main, xlab, ylab) {
  limits <- range(c(truth, estimate), finite = TRUE)

  plot(
    truth, estimate,
    pch = 19,
    col = rgb(0.1, 0.35, 0.75, 0.55),
    xlim = limits,
    ylim = limits,
    xlab = xlab,
    ylab = ylab,
    main = main
  )

  abline(0, 1, col = "firebrick", lwd = 2, lty = 2)

  legend(
    "topleft",
    legend = sprintf("cor = %.3f", cor(truth, estimate)),
    bty = "n"
  )
}

old_par <- par(mfrow = c(2, K), mar = c(4, 4, 3, 1))

for (k in seq_len(K)) {
  plot_comparison(
    L_true_plot[, k],
    L_hat_plot[, k],
    main = sprintf("L, component %d", k),
    xlab = "True normalized L",
    ylab = "Estimated normalized L"
  )
}

for (k in seq_len(K)) {
  plot_comparison(
    B_true_plot[, k],
    B_hat_plot[, k],
    main = sprintf("f0 * F, component %d", k),
    xlab = "True effective F",
    ylab = "Estimated effective F"
  )
}

par(old_par)
```

### Plot mixture proportions convergence

```{r mixprop-iter, fig.width=15, fig.height=5}
mixprop_tolerance <- 0.01
iterations <- 0:niter

# best_permutation maps:
# true factor k -> estimated factor column best_permutation[k]
#
# Apply one fixed final permutation to the complete trajectory.
pi_trace_aligned <- pi_trace[
  ,
  ,
  best_permutation,
  drop = FALSE
]

dimnames(pi_trace_aligned)$estimated_factor <-
  paste0("k", seq_len(K))

# Confirm that the final trace values equal the aligned pi used
# by the mixture-proportion heatmap.
stopifnot(
  isTRUE(all.equal(
    unname(pi_trace_aligned[niter + 1L, , ]),
    unname(pi_hat),
    tolerance = 1e-10
  ))
)

# First iteration after which all proportions remain within the
# tolerance of their final values.
stabilization_iteration <- vapply(
  seq_len(K),
  function(k) {
    final_pi <- pi_trace_aligned[niter + 1L, , k]

    distance_from_final <- apply(
      abs(sweep(
        pi_trace_aligned[, , k],
        2,
        final_pi,
        "-"
      )),
      1,
      max
    )

    future_maximum <- rev(cummax(rev(distance_from_final)))
    first_stable <- which(
      future_maximum <= mixprop_tolerance
    )[1]

    if (is.na(first_stable)) {
      NA_integer_
    } else {
      first_stable - 1L
    }
  },
  integer(1)
)

mixture_colors <- setNames(
  hcl.colors(M, "Dark 3"),
  paste0("a = ", settings$a)
)

old_par <- par(
  mfrow = c(1, K),
  mar = c(4, 4, 3, 1)
)

for (k in seq_len(K)) {
  matplot(
    iterations,
    pi_trace_aligned[, , k],
    type = "l",
    lty = 1,
    lwd = 2,
    col = mixture_colors,
    ylim = c(0, 1),
    xlab = "Iteration",
    ylab = "Mixture proportion",
    main = paste("True factor", k)
  )

  # Generating mixture proportions.
  for (m in seq_len(M)) {
    abline(
      h = synthetic$pi[m, k],
      col = mixture_colors[m],
      lty = 2
    )
  }

  # Empirical stabilization iteration.
  if (!is.na(stabilization_iteration[k])) {
    abline(
      v = stabilization_iteration[k],
      col = "grey40",
      lty = 3
    )
  }

  legend(
    "right",
    legend = names(mixture_colors),
    col = mixture_colors,
    lty = 1,
    lwd = 2,
    cex = 0.85,
    bty = "n"
  )
}

par(old_par)

final_mixprop <- expand.grid(
  mixture = paste0("a=", settings$a),
  factor = paste0("k", seq_len(K))
)

final_mixprop$generating <- as.vector(synthetic$pi)
final_mixprop$estimated <- as.vector(
  pi_trace_aligned[niter + 1L, , ]
)

list(
  alignment = data.frame(
    true_factor = paste0("k", seq_len(K)),
    estimated_column = best_permutation
  ),
  stabilization = data.frame(
    factor = paste0("k", seq_len(K)),
    iteration = stabilization_iteration
  ),
  final_proportions = final_mixprop
)
```

Solid lines show the estimated mixture proportions recorded during the main CAVI run, while matching dashed lines show the generating proportions.

### Plot reconstructed Poisson rate

```{r rate-lineplot}
lambda_true <- L_true %*% t(B_true)
lambda_hat  <- L_hat  %*% t(B_hat)

lambda_limits <- range(c(lambda_true, lambda_hat))

plot(
  as.vector(lambda_true),
  as.vector(lambda_hat),
  pch = 19,
  cex = 0.7,
  col = rgb(0, 0, 0, 0.25),
  xlim = lambda_limits,
  ylim = lambda_limits,
  xlab = "True Poisson mean",
  ylab = "Estimated Poisson mean",
  main = "Reconstructed mean: true versus fitted"
)
abline(0, 1, col = "firebrick", lwd = 2)

lambda_metrics <- data.frame(
  correlation = cor(as.vector(lambda_true), as.vector(lambda_hat)),
  relative_error = sqrt(sum((lambda_hat - lambda_true)^2)) /
    sqrt(sum(lambda_true^2)),
  rmse = sqrt(mean((lambda_hat - lambda_true)^2))
)

lambda_metrics
```

The reconstructed-rate plot compares
\(\lambda=L(f_0F)^\mathsf{T}\), the quantity that directly determines the
Poisson likelihood. Note strong rate recovery is possible even when the individual factors
are not recovered exactly.

### Background recovery

```{r bg-lineplot, fig.width=12, fig.height=5}
loading_scale <- c(
  true = mean(rowSums(L_true)),
  estimated = mean(rowSums(L_hat))
)

if (any(abs(loading_scale - 1) > 1e-2)) {
  warning(sprintf(
    "Loading scales are not both within 1e-2 of 1 (true = %.4f, estimated = %.4f); raw f0 values may not be directly comparable.",
    loading_scale["true"], loading_scale["estimated"]
  ))
}

log_f0_true <- log(f0_true)
log_f0_hat <- log(f0_hat)
plot_limits <- range(log_f0_true, log_f0_hat)
feature <- seq_along(f0_true)

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

matplot(
  feature,
  cbind(log_f0_true, log_f0_hat),
  type = "l",
  lty = c(1, 2),
  lwd = 2,
  col = c("black", "steelblue"),
  xlab = "Feature",
  ylab = "log(f0)",
  main = "Background profile"
)
legend(
  "topright",
  legend = c("True", "Estimated"),
  col = c("black", "steelblue"),
  lty = c(1, 2),
  lwd = 2,
  bty = "n"
)

plot(
  log_f0_true,
  log_f0_hat,
  pch = 19,
  col = rgb(0.1, 0.35, 0.75, 0.55),
  xlim = plot_limits,
  ylim = plot_limits,
  xlab = "True log(f0)",
  ylab = "Estimated log(f0)",
  main = "Background recovery"
)
abline(0, 1, col = "firebrick", lwd = 2, lty = 2)

par(old_par)

log_error <- log_f0_hat - log_f0_true
background_metrics <- data.frame(
  correlation = cor(f0_true, f0_hat),
  log_correlation = cor(log_f0_true, log_f0_hat),
  log_rmse = sqrt(mean(log_error^2)),
  median_fold_error = exp(median(abs(log_error)))
)

background_metrics
```

The plots compare the raw true and estimated `f0` values.

### Plot true vs. estimated loadings, factors

```{r lf-heatmap, fig.width=16, fig.height=18, out.width="100%", dpi=150}
draw_heatmap_wide <- function(
    x,
    main,
    zlim,
    row_labels = NULL,
    xlab = "Index"
) {
  x_plot <- log1p(x)

  image(
    x = seq_len(ncol(x_plot)),
    y = seq_len(nrow(x_plot)),
    z = t(x_plot),
    zlim = zlim,
    col = hcl.colors(100, "Viridis"),
    useRaster = TRUE,
    axes = FALSE,
    xlab = xlab,
    ylab = "",
    main = main
  )

  # Show a few horizontal index labels.
  x_ticks <- unique(round(seq(
    1,
    ncol(x_plot),
    length.out = min(6, ncol(x_plot))
  )))

  axis(1, at = x_ticks, labels = x_ticks)

  if (is.null(row_labels)) {
    row_labels <- seq_len(nrow(x_plot))
  }

  axis(
    2,
    at = seq_len(nrow(x_plot)),
    labels = row_labels,
    las = 1
  )

  box()
}

draw_pair_wide <- function(
    truth,
    estimate,
    label,
    row_labels = NULL,
    xlab = "Index"
) {
  pair_range <- range(
    c(log1p(truth), log1p(estimate)),
    finite = TRUE
  )

  if (diff(pair_range) == 0) {
    pair_range <- pair_range + c(-1, 1) * 1e-8
  }

  draw_heatmap_wide(
    truth,
    main = paste("True", label, ", Max value ", round(max(truth),1)),
    zlim = pair_range,
    row_labels = row_labels,
    xlab = xlab
  )

  draw_heatmap_wide(
    estimate,
    main = paste("Estimated", label, ", Max value ", round(max(estimate),1)),
    zlim = pair_range,
    row_labels = row_labels,
    xlab = xlab
  )
}

factor_labels <- paste0("k", seq_len(K))

old_par <- par(
  mfrow = c(3, 2),
  # With the 16 x 18 inch device above, every heatmap receives a
  # plotting region of roughly 8 x 6 inches instead of being squeezed
  # into one quarter of R Markdown's default-height graphics device.
  mar = c(2.5, 3, 2, 0.5),
  mgp = c(2.4, 0.7, 0),
  tcl = -0.25,
  cex.axis = 0.9,
  cex.lab = 1,
  cex.main = 1.05
)

# K rows × n columns
draw_pair_wide(
  t(L_true_plot),
  t(L_hat_plot),
  label = "L",
  row_labels = factor_labels,
  xlab = "Sample"
)

# K rows × p columns
draw_pair_wide(
  t(B_true_plot),
  t(B_hat_plot),
  label = "f0 * F",
  row_labels = factor_labels,
  xlab = "Feature"
)

par(old_par)
```

Heatmaps use `log1p` values (because L, f0*F likely contain most entries close to zero and a few much larger values). The effective factor \(f_0F\) is emphasized
because it is the feature-side quantity that enters the Poisson rate. However, note separate
differences in `f0` and `F` can compensate for one another.

### Plot true vs. estimated mixture proportions

```{r mixprop-heatmap, fig.width=12, fig.height=5.5, out.width="100%", dpi=150}
draw_mixture_proportions <- function(pi, main, a = NULL) {
  M <- nrow(pi)
  K <- ncol(pi)

  image(
    x = seq_len(K),
    y = seq_len(M),
    z = t(pi),
    zlim = c(0, 1),
    col = hcl.colors(100, "Blues 3", rev = TRUE),
    axes = FALSE,
    xlab = "Factor",
    ylab = "",
    main = main,
    useRaster = TRUE
  )

  axis(
    1,
    at = seq_len(K),
    labels = paste0("k", seq_len(K))
  )

  mixture_labels <- paste0("m", seq_len(M))
  if (!is.null(a)) {
    mixture_labels <- sprintf(
      "m%d (a = %g)",
      seq_len(M),
      a
    )
  }

  axis(
    2,
    at = seq_len(M),
    labels = mixture_labels,
    las = 1
  )

  cells <- expand.grid(
    factor = seq_len(K),
    mixture = seq_len(M)
  )
  values <- as.vector(t(pi))

  text(
    x = cells$factor,
    y = cells$mixture,
    labels = sprintf("%.3f", values),
    col = ifelse(values >= 0.55, "white", "black"),
    font = 2,
    cex = 1.15
  )

  box()
}

old_par <- par(
  mfrow = c(1, 2),
  mar = c(3, 4, 2, 0.5),
  mgp = c(2.4, 0.7, 0),
  tcl = -0.25
)

draw_mixture_proportions(
  pi_true,
  main = "True mixture proportions",
  a = settings$a
)

draw_mixture_proportions(
  pi_hat,
  main = "Estimated mixture proportions",
  a = settings$a
)

par(old_par)
```


The annotated cells report \(\pi_{mk}\) directly. Columns of the estimated
matrix use the same factor matching as the preceding plots, while rows
correspond to the Gamma components identified by their shape parameters
`settings$a`.

### Plot if f0 error is correlated with f_jk error

```{r error-compensate, eval=!oracle_f0}
# If correlation high, error in f0 is offsetting error in F
F_true_canonical <- sweep(B_true_plot, 1, f0_true, "/")
F_hat_canonical  <- sweep(B_hat_plot, 1, f0_hat, "/")

log_f0_error <- log(f0_hat / f0_true)
mean_log_F_error <- rowMeans(
  log(F_hat_canonical / F_true_canonical)
)

plot(
  log_f0_error,
  -mean_log_F_error,
  pch = 19,
  xlab = "log(f0_hat / f0_true)",
  ylab = "-mean log(F_hat / F_true)",
  main = "Compensation between f0 and F"
)
abline(0, 1, col = "firebrick", lwd = 2)

cor(log_f0_error, -mean_log_F_error)
```

If correlation high, error in f0 is offsetting error in F. On the other hand, if the plot is concentrated near the red vertical line at zero, it means little compensation.
