---
title: "EM"
author: "Jihee You"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# install.packages("here")
library(here)
here()
```

# PNMF Problem
\section{Poisson NMF}

Let $X \in R_{+}^{n \times p}$ denote a matrix of observed counts
$x_{ij}$. Poisson NMF parameters include $L \in R_{+}^{n \times k}$
and $F \in R_{+}^{p \times k}$, and we seek values of L and F such
that $X \approx LF^T$.

Given $k \ge 1$, the Poisson NMF model is 
$$
\begin{aligned}
X_{ij} | L, F  \sim Poi(\lambda_{ij}) \\
\lambda_{ij} = \sum_{k} l_{ik}f_{jk}
\end{aligned}
$$

Likelihood for a single observed count $X_{ij}$ is expressed as 
$$
P(X_{ij} | L, F) = \frac{\exp(-\lambda_{ij}) \lambda_{ij}^{X_{ij}}}{X_{ij}!}
$$

Then, the log-likelihood of X is expressed as

$$
\begin{align*}
    \ell(X|L,F) &= \log(\prod_{i}\prod_{j} P(X_{ij} | L, F) \\
    &= \sum_{i}\sum_{j} \log(\frac{\exp(-\lambda_{ij}) \lambda_{ij}^{X_{ij}}}{X_{ij}!}) \\
    &= \sum_{i} \sum_{j}[-\lambda_{ij} + X_{ij}\log(\lambda_{ij}) - \log(X_{ij}!)]
\end{align*}
$$

\subsection{Latent variable representation and EM}

We are able to write the following latent variable representation.
$$
Z_{ijk} \sim Poi(l_{ik} f_{jk}) \\
X_{ij} = \sum_k Z_{ijk}
$$

Let $\hat z_{ijk} = \mathbb{E}[z_{ijk} | x_{ij}, l_{ik}, f_{jk}]$. $\hat z_{ijk}$ can directly calculated.
$$\hat z_{ijk} = x_{ij} \frac{l_{ik}f_{jk}}
{\sum_k {l_{ik}f_{jk}}}$$

The posterior distribution of the latent variable Z is:
$$ P(Z| X, L,F) = \sum_{ijk} (\hat z_{ijk}\log(l_{ik}f_{jk}) - l_{ik}f_{jk} - \log\hat z_{ijk}!)
$$
\[\frac{\partial P}{\partial l_{ik}} = \sum_j (\frac{\hat z_{ijk}}{l_{ik}} - f_{jk})\]
$\frac{\partial P}{\partial l_{ik}} = 0$ leads to
$$
l_{ik}^\text{(new)} = \sum_{j} \frac{\hat z_{ijk}}{f_{jk}}
$$

Similar multiplicative update is available for $f_{jk}$.
$$
f_{jk}^\text{(new)} = \sum_{i} \frac{\hat z_{ijk}}{l_{ik}}
$$


# Data Generation

This step is identical to the main experiment.

```{r}
source(here("docs","code","data_generate.R"))
source(here("docs","code","utils.R"))
library(pheatmap)
library(ggplot2)

# Generate synthetic L, F, X

# Define mixture weights for each L_k and F_k
# 21st mixture component (shape, rate) value is 1
l1 <- c(1, rep(0, 33))
l2 <- c(rep(0, 20),1, rep(0,13))
l3 <- c(rep(0, 33), 1)
f1 <- c(rep(1, 34))
f2 <- c(1, 1, rep(0, 30), 1, 1)
f3 <- c(rep(0, 19), 1, 1, 1, 1, rep(0, 11))

# Weights will be normalized by column inside visualize_mixtures or generate_synthetic
w_matrix <- cbind(l1,l2,l3,f1,f2,f3)

visualize_mixtures(w_matrix)
# "mix", "clear membership"
data <- generate_synthetic(data_type="mix", w_matrix=w_matrix, n=12, p=30, K=3)

# Visualize Ground Truth L, F
visualize_lf(data$L, data$F, "True")
```

# Define likelihood-related functions

## 1. Incomplete data log likelihood

<!-- Eq 11.2 in Jordan chapter -->
Incomplete data log likelihood is defined as 
$$
l(\theta ; x) = \log p(x|\theta) = \log \sum_z p(x, z | \theta)
$$
In this simulation case, the incomplete log likelihood is
<!-- Notice I go from x to X matrix notation 
and i didn't specify L n k matrix etc k popped up out of nowhere
ADD note explaining its a summation of of Poisson PMF (X parametrized by L, F)-->

$$
\begin{align}
l(L,F; X) &= \log p(X| L,F) \\
&= \sum_{i} \sum_{j}[-\lambda_{ij} + X_{ij}\log(\lambda_{ij}) - \log(X_{ij}!)]
\end{align}
$$
given that $\lambda_{ij} = \sum_k L_{ik} F_{jk}$ for simpler notation.

## Implementation: Likelihood
```{r}
library(Matrix)
data$X <- Matrix(data$X, sparse = TRUE)

```

```{r}
library(matrixStats)
source(here("docs","code","utils.R"))

cat("Log likelihood",compute_loglik(log(data$L), log(data$F), data$X), "\n")

# Shuffle L around rows, see how much log likelihood decreases (-569)
#shuffled_L <- data$L[sample(nrow(data$L)),]
#compute_loglik(log(shuffled_L), log(data$F), data$X)
```


## 2. Complete data log likelihood
<!-- Eq 11.1 in Jordan chapter -->
Complete data log likelihood is defined as below. In this case, theta is L and F.
$$
l_c(\theta; x,z) \triangleq \log p (x,z | \theta)
$$

Because the following assumption defines latent variable Z, access to X, L, F determines Z, without any randomness.

$$
Z_{ijk} \sim Poi(l_{ik} f_{jk}) \\
X_{ij} = \sum_k Z_{ijk}
$$

Thus complete data likelihood is equal to incomplete data likelihood within our model formulation.

$$
\begin{align}
l(L,F; X, Z) &= \log p(X, Z| L,F) \\
&= \log [p (Z | X, L, F) \times p(X | L, F)] \\
&= \sum_{i} \sum_{j}1 * [-\lambda_{ij} + X_{ij}\log(\lambda_{ij}) - \log(X_{ij}!)]
\end{align}
$$

## 3. Auxiliary function 
Also known as the lower bound for incomplete likelihood, auxiliary function for EM algorithm is defined as

$$
\mathcal{L}(q,\theta) = \sum_z q(z|x)\log \frac{p(x,z|\theta)}{q(z|x)}
$$

## E-step
Because of the way latent variable Z was defined, we can assume a multinominal distribution modelling z given x. Therefore, (also since multinominal parameter n is fixed to X) updating $q(z|x)$ is just a matter of choosing multinominal probabilities. Also, we know that the auxiliary function is maximized when $q^{(t+1)} (z|x) = p(z| x, \theta^t)$. Therefore, the E-step is

$$
\begin{align}
q^{(t+1)}(Z|X) &= p(z| x, \theta^t) \\
&= MN(n = X, p = \frac{l_{i:}f_{j:}}{\lambda_{ij}})
\end{align}
$$


## M-step
Once $q^{(t+1)}(Z|X)$ is updated, we want 

$$
\theta^{(t+1)} = \operatorname*{argmax}_\theta \mathcal{L}(q^{(t+1)}, \theta) 
$$

<!-- Refer to 11.13 for connection with expected complete log likelihood -->

$$
\begin{align}
\operatorname*{argmax}_\theta \mathcal{L}(q^{(t+1)}, \theta) &= \operatorname*{argmax}_\theta \langle \mathcal{l}_c(\theta ; x,z) \rangle_q \\
&= \operatorname*{argmax}_\theta \sum_z q(z|x) \log p (x,z|\theta) \\
&= \sum_z q(z|x) \operatorname*{argmax}_\theta \log p (x,z|\theta) \\
&= \operatorname*{argmax}_\theta \log p(\mathbb{E}[z] | \theta)
\end{align}
$$
Note $\mathbb{E}[z]$ is given from the multinomial distribution in E-step: $\mathbb{E}[z_{ij}] = X_{ij} \frac{l_{i:}f_{j:}}{\lambda_{ij}}$.
<!-- length k vectors-->

Going back to the latent variable representation $Z_{ijk} \sim Poi(l_{ik} f_{jk})$, $\mathbb{E}[z] \sim Poi(\theta)$ thus we can write  

<!-- Log Poisson PMF for X \sim Poi(\theta) -->

$$
p(\mathbb{E}[z] | \theta) = \sum_{ijk} (\mathbb{E}[z_{ijk}]\log(l_{ik}f_{jk}) - l_{ik}f_{jk} - \log \mathbb{E}[z_{ijk}]!)
$$


Taking partial derivative of log Poisson PMF with respect to $l_{ik}$ leads to 

$$
\begin{align}
\frac{\partial \ p(\mathbb{E}[z] | \theta)}{\partial \ l_{ik}} &= \sum_j (\frac{\mathbb{E}[z_{ijk}]}{l_{ik}} - f_{jk}) \\
\Rightarrow l_{ik}^\text{(new)} &= \sum_{j} \frac{\mathbb{E}[z_{ijk}]}{f_{jk}}
\end{align}
$$

Similarly, 

$$
f_{jk}^\text{(new)} = \sum_{i} \frac{\mathbb{E}[z_{ijk}]}{l_{ik}}
$$

# Implementation: EM

```{r}
# Computes E_Z, which is a n x p x k matrix
E_step <- function(log_L, log_F, X){
  log_Lambda = compute_log_Lambda_mat(log_L, log_F, X)
  loglam_val <- log_Lambda@x
  
  X_summary <-summary(X)
  x_val <- X_summary$x
  i <- X_summary$i
  j <- X_summary$j
  
  # Compute W_ij = X_ij / Lambda_ij (Only for non-zeros)
  W_val <- exp(log(x_val) - loglam_val) # does index correspond?
  W <- sparseMatrix(i=i, j=j, x=W_val, dims=dim(X))
  
  # Fast sparse matrix multiplications in original space
  L <- exp(log_L)
  F <- exp(log_F)
  W_dot_F <- as.matrix(W %*% F)      # n x k
  Wt_dot_L <- as.matrix(t(W) %*% L)  # p x k
  
  # Back to log space
  logEz_sj = log_L + log(W_dot_F)  # Ez with j dimension summed, n x k
  logEz_si = log_F + log(Wt_dot_L) # Ez with i dimension summed, p x k
  
  # Return two squished slices of logEz
  return(list(logEz_sj = logEz_sj, logEz_si = logEz_si))
}

# Update log_L, log_F 
M_step <- function(log_L, log_F, E_res){
  # Prepare inputs
  logEz_sj = E_res$logEz_sj
  logEz_si = E_res$logEz_si
  
  # Update L first
  F <- exp(log_F)
  logF_sj = log(colSums(F)) # length k
  newlog_L = logEz_sj - rep(logF_sj, each = nrow(log_L))
  
  # Then update F
  # L <- exp(log_L)
  L <- exp(newlog_L)
  logL_si = log(colSums(L)) # length k
  newlog_F = logEz_si - rep(logL_si, each = nrow(log_F))
  
  return(list(log_L=newlog_L, log_F=newlog_F))
}
```

```{r}
library(fastTopics)
library(NNLM)
source(here("docs","code","initialize.R"))
source(here("docs","code","utils.R"))

set.seed(4)

init_res = NNLM::nnmf(A = as.matrix(data$X), k = 3)
log_L = log(init_res$W)
log_F = log(t(init_res$H))

# init_res = initialize(X=data$X, K=3)
# log_L = init_res$qg$qllog_norm
# log_F = init_res$qg$qflog_norm

init_ll = compute_loglik(log_L, log_F, data$X)
gt_ll = compute_loglik(log(data$L), log(data$F), data$X)

visualize_lf(exp(log_L),
             exp(log_F),
             sprintf("Initialized Log-Likelihood: %.2f\n", init_ll))

# qg = list(ql_norm = L, qllog_norm = log(L), kl_l = replicate(K,NA),
#             qf_norm = F, qflog_norm = log(F), kl_f = replicate(K,NA),

maxiter=50
ll_list = rep(numeric(maxiter))
for (niter in 1:maxiter){
  Ez_two = E_step(log_L, log_F, data$X)
  new = M_step(log_L, log_F, Ez_two)
  log_L = new$log_L
  log_F = new$log_F

  ll = compute_loglik(log_L, log_F, data$X)
  ll_list[niter] = ll
  # cat("niter ",niter, "-------------", sprintf("LL: %.2f\n",  ll))
}
visualize_lf(exp(log_L), exp(log_F), "End results")

# 1. Calculate y-axis limits so init_ll and gt_ll don't get cut off
y_bounds <- range(c(init_ll, ll_list, gt_ll), na.rm = TRUE)

# 2. Plot the main EM iterations (x = 1 to maxiter)
plot(1:length(ll_list), ll_list, 
     xlim = c(0, length(ll_list)),  # Expands x-axis to include 0
     ylim = y_bounds,               # Expands y-axis to fit all values
     main = sprintf("Log Likelihood over EM Iterations\nTrue LL: %.2f", gt_ll),
     ylab = "Log-Likelihood",
     xlab = "EM Iteration",
     type = "b",                    # 'b' draws both points and lines
     pch = 16)                      # Solid circle marker

# 4. Plot the initial LL point at x = 0 in red (cex = 1.5 makes it slightly larger)
points(0, init_ll, col = "red", pch = 16, cex = 1.5)

# 5. Add the True LL as a horizontal dashed line
abline(h = gt_ll, col = "blue", lty = 2, lwd = 2)

```

# Evaluate
```{r}
source(here("docs","code","utils.R"))
source(here("docs","code","eval.R"))

mae = eval_mae(est_L=exp(log_L), est_F=exp(log_F), 
               true_L=data$L, true_F=data$F, 
               X=data$X)

csim = eval_csim(est_L=exp(log_L), est_F=exp(log_F), 
                 true_L=data$L, true_F=data$F)

plot_upper_heatmap(csim$cos_L, "L Cosine Similarity")
plot_upper_heatmap(csim$cos_F, "F Cosine Similarity")

```

# Explanations for upkeep, irrelevant to implementation though

Observed data $X \in \mathbb{R}^{n \times p}$ may have intricate internal dependencies that we don't know how to model. But introducing  but assuming a latent Z exists

$\log p(x, z| \theta) $ is the complete log likelihood, can be obtained if latent Z was observed. $Z= LF^T$  in my case is the clean data where the real world data was pulled from with $X = Poi (Z)$. However, Z cannot be observed, so we are forced to work with incomplete log likelihood, equal to marginalized complete log-likelihood.
$$ 
\begin{align}
\log p(x|\theta) &= log \sum_z p(x,z | \theta) \\
&= l(\theta ; x) \text{<-  log likelihood}
\end{align}
$$

Given that Z is not observed, complete log likelihood $\log p(x, z| \theta)$ is a random quantity. To remove the randomness in complete log likelihood, we can average over z drawn from a estimated distribution $q(z|x,\theta). Then we can define the expected complete log likelihood as
$$
\langle l(\theta ; x, z) \rangle _q = \sum_z q (z | x,\theta) \log p (x,z | \theta)
$$
Note posterior probability of the latent variable z $p(z| x, \theta^{(t)})$ is the optimal $q^{(t+1)}(z|x)$ with 

Averaging over z removes stochasticity associated with z, making expected complete log likelihood no longer a random variable but a deterministic function over $\theta$. This ensures auxiliary function $L(q, \theta^{t})$ (also the lower bound on log likelihood) becomes equal to log likelihood.

