---
title: "simulation and omim demo"
author: "Jihee You"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
here()
```

## Simulation, Visualization - Related Imports
```{r}
source(here("docs","code","load_data.R"))
source(here("docs","code","utils.R"))
library(Matrix)
set.seed(1)

source(here("docs","code","viz.R"))
library(pheatmap)
library(ggplot2)
```

## Main Code Implementation for Model A, M=1 Case

### Explanation for ebpmf_modelA_m1 steps 1-3

Similar to E-step in EM, we derived from "Update Z" section the following result:

$$
\mathbb{E}_q[z_{ijk}] = \frac{x_{ij}}{\sum_{k=1}^K h_{ik}w_{jk}} h_{ik} w_{jk}
$$
Let $A_{ij}$ denote $\frac{x_{ij}}{\sum_{k=1}^K h_{ik}w_{jk}}$. Then,

$$
\sum_{j=1}^p \mathbb{E}_q [z_{ijk}] = h_{ik} \sum_{j=1}^p A_{ij} w_{jk}
$$

$$
\sum_{i=1}^n \mathbb{E}_q [z_{ijk}] = w_{jk} \sum_{i=1}^n A_{ij}h_{ik}
$$

### Reference to writeup for ebpmf_modelA_m1 steps 4, 5

Step 4 updates W according to equation 19 in writeup.
$$
w_{jk}^* = \frac{a_k + \sum_{i=1}^n \langle z_{ijk} \rangle}{b_k + \sum_{i=1}^n h_{ik}}
$$

Step 5 updates H according equation 26 in writeup.

$$
\hat h_{ik} = \frac{\sum_{j=1}^p \langle z_{ijk} \rangle}{\sum_{j=1}^p \mathbb{E}_q[w_{jk}]}
$$

### Code

```{r}
ebpmf_modelA_m1 <- function(X, 
                            K=3, 
                            a = c(1e-16, 1, 1000), 
                            b = c(1e-16, 1, 1000), 
                            max_iter = 50) {
  n <- nrow(X)
  p <- ncol(X)
  
  # Initialize H and W with random values [0,1]
  H <- matrix(runif(n * K, 0, 1), nrow = n, ncol = K)
  W <- matrix(runif(p * K, 0, 1), nrow = p, ncol = K)
  
  # Extract sparse matrix components for easy index access when working with H or W
  X_i <- X@i + 1            # 0-indexed -> 1-indexed rows
  X_p <- X@p                # Column pointers
  X_j <- rep(seq_along(X_p[-1]), diff(X_p)) # Reconstruct columns
  X_val <- X@x              # Store nonzero values
  
  for (iter in 1:max_iter) {
    # 1. Compute expected Poisson rates
    # Denoted as lambda_{ij}, but actually \sumk \langle z_{ijk} \rangle 
    # Lambda_ij = \sum_k h_ik w_jk
    Lambda_val <- rowSums(H[X_i, , drop = FALSE] * W[X_j, , drop = FALSE])

    # 2. Compute A_ij = X_ij / Lambda_ij
    # A and X have same dimension: n x p
    A_val <- X_val / (Lambda_val + 1e-20)
    A <- sparseMatrix(i = X_i, j = X_j, x = A_val, dims = c(n, p))
    
    # 3. Compute expected Z (similar to E-step)
    Z_H <- H * as.matrix(A %*% W) # This is \sumj E_q[Z]
    Z_W <- W * as.matrix(crossprod(A, H)) # This is \sumi E_q[Z]
    
    ######
    # Check dimensions
    # print(length(a))
    # print(dim(Z_H)) nk
    # print(dim(Z_W)) pK
    
    # Sweep demonstrations
    # print(Z_W)
    # print(a)
    # print(sweep(Z_W, 2, a, "+"))
    ######

    # 4. Update W (Equation 19)
    sum_H <- colSums(H)
    W_new <- sweep(sweep(Z_W, 2, a, "+"), 2, b + sum_H, "/") # Add k-length 'a' to each row of k entries, divide by (b + sum_H) for each row
    
    # 5. Update H (Equation 26)
    sum_W <- colSums(W_new) 
    H_new <- sweep(Z_H, 2, sum_W, "/")
    
    # Early stop?
    delta <- max(abs(H - H_new), abs(W - W_new))
    if (delta < 1e-6) { cat("Converged at iter", iter, "\n"); break }
    
    # Apply updates
    H <- H_new
    W <- W_new
    }
  
  return(list(H = H, W = W))
}
```

```{r}
# Experiment 1:
# When randomness in X comes from
# 1) Randomness in factors W
#     k=1: W_{jk} \sim Gamma(shape=1e-16, rate = 1/shape)
#     k=2: W_{jk} \sim Gamma(shape=1, rate = 1/shape)
#     k=3: W_{jk} \sim Gamma(shape=1000, rate = 1/shape)
# 2) X \sim Poisson generation process
# NOT randomness in loadings H. H_{ik} = 0.5

data <- general_load(data_type="modelA_unifL")
plot_lf(data$L, data$F, "True")

X <- Matrix(as.matrix(data$X), sparse = TRUE)
res = ebpmf_modelA_m1(X, max_iter=50)
plot_lf(res$H, res$W, "Estimated")
```

```{r}
# Experiment 2:
# Some randomness in loadings H as well, since H_{ik} \sim Gamma(1,1)

data <- general_load(data_type="modelA_gammashape1L")
plot_lf(data$L, data$F, "True")

X <- Matrix(as.matrix(data$X), sparse = TRUE)
res = ebpmf_modelA_m1(X, max_iter=50)
plot_lf(res$H, res$W, "Estimated")
```

## Fasttopics on Omim (Peter's code)

```{r fasttopics}
library(tools)
library(fastTopics)
load(file=here("docs","data", "omim.RData"))
set.seed(1)
k <- 40
tm <- fit_poisson_nmf(counts,k = k,init.method = "random",method = "em",
                      numiter = 20,verbose = "detailed",
                      control = list(numiter = 4,nc = 4,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 100,
                      control = list(numiter = 4,nc = 4,extrapolate = FALSE),
                      verbose = "detailed")
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 100,
                      control = list(numiter = 4,nc = 4,extrapolate = TRUE),
                      verbose = "detailed")
save(list = "tm",file = "omim_topics_k=40.RData")
resaveRdaFiles("omim_topics_k=40.RData")

# Inspecting the topics:
k <- 1
L <- poisson2multinom(tm)$L
F <- poisson2multinom(tm)$F
i <- head(order(L[,k],decreasing = TRUE),n = 20)
j <- which(F[,k] > apply(F[,-k],1,max))
j <- head(j[order(F[j,k],decreasing = TRUE)],n = 30)
keywords <- rownames(F)[j]
dat <- data.frame(L     = L[,k],
                  mim   = meta_data$mim,
                  title = substr(meta_data$title,1,60))
cat("Topic ",k,":\n",sep = "")
print(format(dat[i,],justify = "left"),
      row.names = FALSE)
cat("keywords:\n")
print(keywords,quote = FALSE)
```

